### 2017-07-29 KineticModel/estimation/docs/formulation/noisy_estimate.tex
copy identifiability analysis text from q document from 2013

>>>>

\paragraph{Identifiability Analysis:}
Parameter uncertainties accompanying parameter inference and estimation in traditional deterministic models can also be reduced. Parameter uncertainties in deterministic models can be reduced by making sure that parameters in a given model are structurally (identifiability) and practically (estimability) identifiable (Nikerel et al., 2006). It has been shown that reducing the number of both structurally and practically non-identifiable parameters can result in reduction of model uncertainty propagated through uncertain parameter estimates (Hengl et al., 2007).
The effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter. Assume a dynamic model for a model state variable vector x is represented by Equation 2.5 where u is the vector of input trajectories to the model and p is the vector of model parameters, and the observables y obtained from experimental data are represented in Equation 2.6 where ε is the intrinsic noise associated with the data. If elements of p are estimated based on the noisy data y, then a parameter pi in p is structurally identifiable if and only if there is a unique solution to the system in Equations 2.5-2.6 (Hengl et al., 2007). If many solutions exist for pi, then the parameter is said to be locally identifiable. Consequently, absence of any solution would result in the parameter being termed as non-identifiable in the context of Equations 2.5 and 2.6.

Structural identifiability can also be interpreted in an alternative way based on maximum likelihood estimates (MLE) with respect to a given parameter (Raue, et al., 2009). A parameter is non-identifiable if a unique minimum, to the maximum likelihood estimate with respect to that parameter does not exist. Hence, parameters cannot be truly inferred if they are structurally non-identifiable. Several techniques used to differentiate structurally identifiable parameters from non-identifiable parameters were reviewed by McLean \& McAuley (2012). The advantages and disadvantages of using these techniques for highly nonlinear models that are frequently encountered in biology were discussed therein as well.
In contrast to structural parameter identifiability, practical identifiability of a parameter helps in determining the ability and effects of the available experimental data on estimating model parameter values uniquely. Practical identifiability of a parameter is contingent upon the nature, quality and quantity of data available to estimate the parameter as opposed to the structure and parameterization of the model. It can be gathered that both the model and the data play an important role in the ability to estimate parameters of a model. However, structural identifiability of any parameter in a given model depends on the fundamental biological process being modeled. Modifications to model structure to alleviate structural identifiability should not upend the purpose of the model, i.e. to represent the biological process and the underlying dynamics. Hence improvements in this regard are few and far between as more biological knowledge becomes available. Conversely, the ability to improve confidence in parameter estimates by decreasing practical non-identifiability of model parameters is a convenient solution. 
This can be done by improving the quality and quantity of data obtained for parameter estimation. Experimental design (ED) makes it possible to predetermine the quality and quantity of measurements obtained from experiments for the purpose of parameter estimation. The Fisher information matrix (FIM)-based ED strategies (McLean \& McAuley 2012) are one of the many ED methodologies out there. Most if not all are based on senstivity-based measures of the model predicted state variable to each parameter in the model. However, biological models are highly nonlinear and are subject to high-dimensional parameterization. As a result, increase in network size for large models (genome-scale models for example) increases the computational complexity of calculating such sensitivity matrices required for ED. Techniques based on graphical analysis of the sensitivity curves are also rendered moot due to the high-dimensional parameter space (McLean \& McAuley 2012).
Profile likelihood (PL) based ED for parameter estimation aims to overcome this shortcoming in order to reduce practical as well as structural parameter identifiability. A complete description of PL based ED is provided in Appendix C. Although PL has been applied to small (4-12 parameters) and medium scale (15-40 parameters) networks (Raue, et al., 2009; Steiert et al., 2012), an application to a large scale network has not been accomplished. In addition, currently there is an absence of an ED-based strategy for estimating parameters for large scale transcriptional regulatory and metabolic network models.

PL of any parameter θi,   in a vector θ of parameters is a measure of the log likelihood parameter estimate (LLE) as defined by Equation C.1. This methodology proposed by Raue et al. (2009) determines the PL of a parameter by minimizing the LLE of a parameter with respect to all other parameters in the model. The most important deliverable from this technique is the ability to distinguish between practically identifiable and non-identifiable parameters through confidence interval estimates for the said parameter. Confidence intervals are estimated from the PLs according to Equation C.2. The LLE of any given parameter can be estimated using logarithm of Equation C.3 that gives the likelihood estimate for a parameter vector θ based on a known vector of observables Y that are embedded with Gaussian noise.

In Equation C.2, Δα is the α quartile of the chi-squared distribution with df degrees of freedom. Quantities ym and   in Equation C.3 correspond to the measured and predicted value of observable m respectively at time points tj in a time series. The variance of each observable m is written as  . 
Examples of (practically) identifiable and non-identifiable parameters are provided in Figure C.1 adapted from (Raue, et al., 2009). A practically identifiable parameter is distinguished from a non-identifiable one by the nature of the PL curve. In Panel A, the PL curve is asymptotic giving rise to an unbounded confidence interval for the corresponding parameter θi. This essentially would result in a huge parameter space that is not good enough for a confident model. Experimental data capable of distinguishing observables based on changes in a given parameter are simulated (Panel B of Figure C.1) and used to calculate a fresh PL for the said parameter in Panel A. Accordingly, the nature of the PL curve could change from being asymptotic to quadratic as shown in Panel C. This gives a very narrow and well-defined confidence interval for the estimated parameter unlike the case demonstrated in Panel A. Thus the nature of experiment to be performed and data to be collected is gleaned at the data simulation stage. Experimental data obtained through these designed experiments have been shown to change the characteristic of PL curves in small networks of ~40 parameters (Steiert et al., 2012). It is postulated that this way parameter uncertainties can be reduced and more confidence can be infused into model predictions. In addition to its utility as an experimental design aid, the algorithmic approach followed for PL estimation enables its integration into the computational model reconstruction and validation paradigm being developed as part of the current proposal.
